---
title: "India’s RAMSAR Sites: A Geometric Analysis"
subtitle: "Evaluating Wetland Conservation Trends and Ecological Data"
author: "Shashank Venkatesh"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

## Data Source

The foundational data for this analysis is sourced from the **Open Government Data (OGD) Platform India** catalog on **Ramsar Sites Details**.
This dataset provides the geometric coordinates, area in hectares, and designation dates for the wetlands, allowing for a precise spatial and temporal analysis.

```{r}
#| label: load the dataset
library(jsonlite)
library(dplyr)

# URL of the RAMSAR API
api_url <- "https://indianwetlands.in/wp-content/themes/wetlands/restapi-ramsar.php"

# Fetch and parse the data
# Note: Sometimes government APIs require a 'User-Agent' header to allow access
ramsar_json <- fromJSON(api_url)

# Convert to a data frame
# The API usually returns a nested list; we simplify it here
ramsar_data <- as.data.frame(ramsar_json)

```

## Glimpse of the dataframe

```{r}
#| label: glimpse into the dataframe
#| message: false
#| warning: false
glimpse(ramsar_data)
```

The dataset has **`r nrow(ramsar_data)`** rows and **`r ncol(ramsar_data)`** columns.
