---
title: "India’s RAMSAR Sites: A Geometric Analysis"
subtitle: "Evaluating Wetland Conservation Trends and Ecological Data"
author: "Shashank Venkatesh"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

## Data Source

The foundational data for this analysis is sourced from the **Open Government Data (OGD) Platform India** catalog on **Ramsar Sites Details**.
This dataset provides the geometric coordinates, area in hectares, and designation dates for the wetlands, allowing for a precise spatial and temporal analysis.
#

```{r}
#| label: load the dataset
library(tidyverse)
library(tidymodels)
library(scales) # For formatting legend numbers
library(jsonlite)
library(sf)
#library(rnaturalearth) # Error in Indian map with wrong administrative boundaries
#library(rnaturalearthdata) #
library(mapindia)
library(ggrepel) # For non-overlapping labels
library(plotly)
library(tidytext)
library(patchwork) # To combine plots if needed

# URL of the RAMSAR API
api_url <- "https://indianwetlands.in/wp-content/themes/wetlands/restapi-ramsar.php"

# Fetch and parse the data
# Note: Sometimes government APIs require a 'User-Agent' header to allow access
ramsar_json <- fromJSON(api_url)

# Convert to a data frame
# The API usually returns a nested list; we simplify it here
ramsar_data <- as.data.frame(ramsar_json)

```

## Glimpse of the dataframe

```{r}
#| label: glimpse into the dataframe
#| message: false
#| warning: false
glimpse(ramsar_data)
```

The dataset has **`r nrow(ramsar_data)`** rows and **`r ncol(ramsar_data)`** columns.

## Data Cleaning and Preparation

to ensure all "geometric coordinates" and sizes are correctly typed.

```{r}
#| label: Converting chr strings to proper data types
#| message: false
#| warning: false
# Convert character strings to numeric and date types
ramsar_clean <- ramsar_data |>
  mutate(
    Area = as.numeric(Area),
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    Date = as.Date(DesignationDate),
    Year = as.numeric(format(Date, "%Y"))
  ) |>
  arrange(Date) |>
  mutate(cumulative_sites = row_number())
```

## Geographical Distribution of Wetlands in India

```{r}
#| label: Geographical Distribution of wetlands in India
#| eval: false
#| include: false
# 2. Get India's high-resolution boundary (sf format)
india_map <- ne_countries(scale = "large", country = "India", returnclass = "sf")

# 3. Convert our RAMSAR dataframe into a spatial object (sf)
# We ensure Latitude/Longitude are numeric first
ramsar_sf <- ramsar_clean |>
  filter(!is.na(Latitude) & !is.na(Longitude)) |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 4. Plotting the Geometric Map
ggplot() +
  # Draw the base map of India with a minimalist geometric style
  geom_sf(data = india_map, fill = "#F8F9FA", color = "#ADB5BD", size = 0.3) +
  
  # Plot wetlands as geometric bubbles
  # Size is proportional to the Area; color is our signature Teal
  geom_sf(data = ramsar_sf, 
          aes(size = Area), 
          color = "#2A9D8F", 
          alpha = 0.6) +
  
  # Add labels only for the top 5 largest sites to keep it "geometric and clean"
  geom_text_repel(data = ramsar_sf |> slice_max(Area, n = 5),
                  aes(label = WetlandName, geometry = geometry),
                  stat = "sf_coordinates",
                  size = 3, 
                  fontface = "bold",
                  family = "sans",
                  color = "#264653",
                  box.padding = 0.5) +
  
  # Formatting the scales and legend
  scale_size_continuous(range = c(1, 12), 
                        labels = label_comma(),
                        name = "Area (Hectares)") +
  
  # Apply a minimalist void theme
  theme_void() +
  labs(
    title = "GEOSPATIAL ARCHITECTURE OF INDIA'S RAMSAR SITES",
    subtitle = "Analysis of 96 internationally significant wetlands as of Dec 2025",
    caption = "Data Source: Open Government Data (OGD) Platform India | Projection: WGS84"
  ) +
  
  # Styling the typography
  theme(
    plot.title = element_text(family = "sans", face = "bold", size = 18, color = "#264653", hjust = 0.5),
    plot.subtitle = element_text(family = "sans", size = 11, color = "#666666", hjust = 0.5, margin = margin(b = 20)),
    legend.position = c(0.85, 0.2),
    plot.background = element_rect(fill = "white", color = NA)
  )

```

```{r}
# 1. Retrieve the official state boundaries as an sf object
# This function is specifically designed to provide the official Indian map
india_official <- map_india(regions = "states")

#Convert our RAMSAR dataframe into a spatial object (sf)
# We ensure Latitude/Longitude are numeric first
ramsar_sf <- ramsar_clean |>
  filter(!is.na(Latitude) & !is.na(Longitude)) |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 2. Plotting the correct geometry
ggplot() +
  # Draw the base map with the official boundaries
  geom_sf(data = india_official, fill = "#F8F9FA", color = "#ADB5BD", size = 0.3) +
  
  # Layer the RAMSAR sites as geometric nodes
  geom_sf(data = ramsar_sf, 
          aes(size = Area), 
          color = "#2A9D8F", 
          alpha = 0.6) +
  
  theme_void() +
  labs(
    title = "OFFICIAL SPATIAL ARCHITECTURE: RAMSAR SITES",
    subtitle = "Analysis using official Survey of India administrative boundaries",
    caption = "Data Source: mapindia package & OGD India"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#264653", hjust = 0.5),
    legend.position = "bottom"
  )
```

The above map illustrates that India's RAMSAR sites are not uniformly distributed but are clustered. We can see a concentration of sites in the **Indo-Gangetic Plain** and a significant concentration along the **Southern Coast (Tamil Nadu)**. The nodal points also indicate that the largest wetlands (by area) are mostly located in the North and East, while the South is characterized by a cluster of smaller sites.

## The Acceleration of Conservation: A Temporal Growth Analysis

### How has the rate of RAMSAR designations in India changed over the last three decades, particularly since the 2020s?

to pinpoint the "inflection point" where India’s wetland conservation policy shifted into high gear.

```{r}
#| label: temporal growth analysis
#| message: false
#| warning: false
ggplot(ramsar_clean, aes(x = Date, y = cumulative_sites)) +
  #geom_area(fill = "#2A9D8F", alpha = 0.2) +
  geom_line(color = "#2A9D8F", size = 0.5) +
  geom_point(data = filter(ramsar_clean, Year >= 2020), 
             color = "#E63946", size = 1) +
  labs(
    title = "The Exponential Rise of Indian RAMSAR Sites",
    subtitle = "Cumulative designations from 1981 to December 2025",
    x = "Year of Designation", y = "Total Number of Sites"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 16, color = "#264653"))
```

It is apparent from the above plot that a **Policy Inflection Point may have occurred around 2020**. While designations were sporadic from 1981 to 2010, the curve steepens sharply in the last five years with more than half (59) of India's current Ramsar sites designated since 2020, signaling an aggressive shift in national environmental policy and a heightened commitment to international conservation standards.

```{r}
#| label: temporal analysis of designation 

# 1. Categorize the sites based on the 2020 inflection point
ramsar_phases <- ramsar_clean |>
  mutate(Phase = ifelse(Year < 2020, "Before 2020", "Since 2020")) |>
  count(Phase) |>
  mutate(Percentage = n / sum(n) * 100)

# 2. Print the summary table to console
#print(ramsar_phases)

# 3. Geometric Vector Visualization
ggplot(ramsar_phases, aes(x = Phase, y = n, fill = Phase)) +
  # Using a fixed width for a clean geometric look
  geom_col(width = 0.6, color = "#264653", size = 1) +
  # Adding count labels directly on top of the bars
  geom_text(aes(label = n), vjust = -0.5, size = 6, fontface = "bold", color = "#264653") + 
  ylim(0, 70) +
  # Applying the Teal and Coral palette
  scale_fill_manual(values = c("Before 2020" = "#2A9D8F", "Since 2020" = "#E63946")) +
  labs(
    title = "PHASE ANALYSIS: THE 2020 CONSERVATION PIVOT",
    subtitle = "Comparing decades of legacy designations vs. recent geometric growth",
    x = NULL, 
    y = "Number of Designated Sites"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16, color = "#264653"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold", color = "#264653")
  )
```

As the plot confirms, before 2020 (The Legacy Era), we only had 37 RAMSAR sites. This represents nearly 40 years of gradual identification and designation since India joined the convention in 1982. However, post-2020 (The Modern Era), we can see that 59 additional sites were designated as RAMSAR sites, which forms **61% of India’s total RAMSAR sites were declared in just the last 5 years.** This sudden surge indicates a shift in environmental priority, where the administrative speed of protecting wetlands has tripled compared to the previous four decades.

## State-wise Distribution: The Pareto of Wetlands

### Does a small number of states hold the majority of India's RAMSAR designations, or is it evenly distributed?

Count sites by `StateName` and sort them in descending order.

Highlights regional leaders in ecological preservation.

```{r}
#| label: State-wise Distribution

ramsar_clean |>
  count(StateName) |>
  mutate(Highlight = ifelse(n >= 10, "Top Tier", "Other")) |>
  ggplot(aes(x = n, y = reorder(StateName, n), fill = Highlight)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("Top Tier" = "#E63946", "Other" = "#2A9D8F")) +
  labs(
    title = "Regional Distribution of Wetlands",
    subtitle = "Tamil Nadu leads the geometric distribution with 20 sites",
    x = "Number of Designated Sites", y = NULL, fill = "Status"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 9, face = "bold"))
```

```{r}
#| label: states with highest wetland sites
#| echo: false

top_states <- ramsar_clean |>
  group_by(StateName) |>
  summarise(TotalSites = n()) |>
  arrange(desc(TotalSites))

top_two <- percent(sum(top_states$TotalSites[1:2])/sum(top_states$TotalSites[1:nrow(top_states)]), accuracy = 0.1)
top_five <- percent(sum(top_states$TotalSites[1:5])/sum(top_states$TotalSites[1:nrow(top_states)]), accuracy = 0.1)


niche_states_data <- ramsar_clean |>
  group_by(StateName) |>
  summarise(SiteCount = n()) |>
  filter(SiteCount %in% c(1, 2)) 

summary_counts <- niche_states_data |>
  group_by(SiteCount) |>
  summarise(NumberOfStates = n())

total_states = nrow(top_states)

one_or_two_states = sum(summary_counts$NumberOfStates[1:2])
one_or_two_states_p = percent(one_or_two_states/total_states, accuracy = 0.1)
```

The bar plot confirms a disparity in wetland distribution. **Tamil Nadu** leads the country with **`r top_states$TotalSites[1]` sites**, followed by Uttar Pradesh (**`r top_states$TotalSites[2]`**). These top two states have **`r top_two`** of the total number of RAMSAR sites in India. Further, the share of the top 5 states in this metric is an overwhelming —**`r top_five`**. Conversely, **`r one_or_two_states_p` of states (`r one_or_two_states`)** have only 1or 2 sites, suggesting that conservation success is currently driven by specific state-level initiatives rather than a perfectly uniform national spread.

## The Size Spectrum: Are we protecting large basins or small habitats?

### What is the distribution of wetland sizes (Area), and are there "mega-wetlands" that skew the national average?

Transform the `Area` column to numeric and create a histogram or density plot.

Shows if India focuses on massive ecological systems or smaller, niche biodiversity hotspots.

```{r}
#| label: Large basins v/s small habitats
ggplot(ramsar_clean, aes(x = Area)) +
  geom_histogram(fill = "#264653", color = "white", bins = 20) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Distribution of Wetland Area (Log Scale)",
    subtitle = "Analysis of size variance from small reserves to massive basins",
    x = "Area in Hectares (Log10)", y = "Frequency of Sites"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Latitude & Area: The North-South Conservation Divide

### Is there a relationship between geographical latitude and the size of designated wetlands?

Helps visualize if northern (Himalayan) wetlands tend to be larger or smaller than southern/peninsular ones.

```{r}
#| label: geographical distribution
ggplot(ramsar_clean, aes(x = Latitude, y = Area)) +
  geom_point(aes(size = Area), color = "#2A9D8F", alpha = 0.5) +
  geom_smooth(method = "lm", color = "#E63946", se = FALSE, linetype = "dashed") +
  scale_y_log10(labels = comma) +
  labs(
    title = "Geospatial Area Trends",
    subtitle = "Correlation between Latitude and Wetland Scale",
    x = "Latitude (Degrees North)", y = "Area (Hectares)"
  ) +
  theme_minimal() +
  guides(size = "none")
```

The Log-scaled histogram reveals that India protects a wide "Size Spectrum." Most sites fall within the **1,000 to 10,000-hectare range**. However, the presence of sites on the extreme right of the log scale indicates "Mega-Wetlands" (massive river basins or lakes) that serve as critical national ecological anchors, while the smaller peaks represent specialized bird sanctuaries or niche habitats.

## District-Level Granularity within Leading States

### Within the states with the most RAMSAR sites, are the wetlands clustered in specific districts or spread across the state?

Identifies "ecological hubs" at a local government level, which is useful for management authority analysis.

```{r}
#| label: Identifying ecological hubs
ramsar_clean |>
  filter(StateName == "Tamil Nadu") |>
  count(districtname) |>
  ggplot(aes(x = n, y = reorder(districtname, n))) +
  geom_segment(aes(x = 0, xend = n, yend = districtname), color = "#2A9D8F") +
  geom_point(size = 4, color = "#E63946") +
  labs(
    title = "Local Concentration: Tamil Nadu",
    subtitle = "District-wise breakdown of sites in India's leading RAMSAR state",
    x = "Number of Sites", y = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

The scatter plot with the regression line shows a fascinating geospatial trend: **Northern wetlands tend to be larger in area.** As latitude increases, we see more "Mega-Wetlands." This is likely due to the vast floodplains of the Himalayan river systems. In contrast, southern latitudes (closer to the equator) feature a higher number of smaller, fragmented coastal and tank ecosystems.

## The "Anthropogenic Shift": Natural vs. Human-Made

This analysis reveals if India is increasingly relying on man-made reservoirs for international conservation status and determines if India’s conservation strategy is shifting toward protecting infrastructure-based biodiversity.

Recent additions like **Siliserh Lake** (Rajasthan) and **Kopra Reservoir** (Chhattisgarh) are artificial ecosystems.

### What percentage of India's RAMSAR network is comprised of human-made "Artificial" wetlands, and how does their average size compare to natural basins?

```{r}
#| label: Artificial wetlands
# 1. Classification Logic
ramsar_cat <- ramsar_clean |>
  mutate(Type = case_when(
    str_detect(WetlandName, "Reservoir|Dam|Barrage|Tank|Canal|Reserve") ~ "Human-Made / Managed",
    TRUE ~ "Natural / Semi-Natural"
  ))

# 2. Visualization
ggplot(ramsar_cat, aes(x = Type, y = Area, fill = Type)) +
  geom_violin(alpha = 0.3, color = NA) +
  geom_boxplot(width = 0.1, color = "#264653", outlier.shape = NA) +
  geom_jitter(aes(color = Type), alpha = 0.5, width = 0.15) +
  scale_y_log10(labels = label_comma()) +
  scale_fill_manual(values = c("Human-Made / Managed" = "#E63946", "Natural / Semi-Natural" = "#2A9D8F")) +
  scale_color_manual(values = c("Human-Made / Managed" = "#E63946", "Natural / Semi-Natural" = "#2A9D8F")) +
  labs(title = "Scale Architecture: Natural vs. Managed Ecosystems",
       subtitle = "Comparison of total area distribution (Hectares) across 96 sites",
       x = NULL, y = "Area (Hectares, Log Scale)") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```

The violin-and-box plot reveals that human-made reservoirs and barrages are now statistically comparable in importance to natural lakes. The **Human-Made / Managed** category shows a dense cluster of designations, proving that India is successfully leveraging its irrigation infrastructure (tanks and reservoirs) to provide international-standard habitats for migratory species.

## Semantic Analysis of "Uniqueness": The Biodiversity Fingerprint

(good for qualitative data analysis (categorical data))

### What are the top three "Ecological Drivers" (e.g., Migratory Birds, Flood Control, Carbon Sequestration) that qualify an Indian site for RAMSAR status?

We extract the "Keywords of Uniqueness" to see what makes a wetland internationally significant in the Indian context.

```{r}
#| label: analyse the uniqueness
# 1. Text Mining
uniqueness_words <- ramsar_clean |>
  unnest_tokens(word, Uniqueness) |>
  anti_join(stop_words) |>
  filter(!word %in% c("wetland", "ramsar", "india", "sites", "lake")) |>
  count(word, sort = TRUE) |>
  head(15)

# 2. Geometric Bar Plot
ggplot(uniqueness_words, aes(x = n, y = reorder(word, n))) +
  geom_col(fill = "#264653") +
  geom_text(aes(label = n), hjust = -0.2, size = 3, color = "#264653") +
  labs(title = "The Vocabulary of Conservation",
       subtitle = "Top recurring terms in site uniqueness descriptions",
       x = "Frequency of Mention", y = NULL) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

The word frequency analysis of "Uniqueness" descriptions identifies **"Migratory," "Birds," and "Shelter"** as the core themes. This suggests that India’s Ramsar identity is primarily defined by the **Central Asian Flyway**. The "Biodiversity Fingerprint" of these sites is less about stagnant water and more about their role as vital transit hubs for global avian populations.

## The "Seasonality" of Global Recognition

### Is there a "Policy Momentum" effect where sites are more likely to be designated around specific international events (e.g., World Wetlands Day in February or Independence Day)?

Reveals if political timelines influence ecological milestones. A radial plot to see if designations cluster around specific months (Policy Momentum).

```{r}
#| label: seasonality-analysis
ramsar_months <- ramsar_clean |>
  mutate(Month = month(Date, label = TRUE)) |>
  count(Month)

ggplot(ramsar_months, aes(x = Month, y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_polar() +
  scale_fill_gradient(low = "#8ECAE6", high = "#219EBC") +
  labs(title = "The Annual Cycle of Designations",
       subtitle = "Frequency of RAMSAR announcements by month",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

The radial chart shows that Ramsar designations are not distributed evenly throughout the year. There is a visible clustering in specific months (often aligning with **World Wetlands Day in February** or year-end policy cycles). This suggests that the "timing" of conservation is often tied to global environmental milestones and diplomatic windows.

## Fragmented vs. Clustered Landscapes (Spatial Join)

### Are Indian RAMSAR sites becoming "Ecological Corridors" (closely clustered) or "Isolated Oases"?

Identify Hotspots (like the Tamil Nadu belt) versus Isolated sites (like Pala Wetland in Mizoram). This calculates how far each site is from its closest "Conservation Neighbor" to identify hubs.

```{r}
#| label: Ecological Corridors
# 1. Calculate distances
coords <- st_coordinates(ramsar_sf)
dist_matrix <- st_distance(ramsar_sf)
diag(dist_matrix) <- Inf # Ignore distance to self
ramsar_clean$min_dist_km <- apply(dist_matrix, 1, min) / 1000

# 2. Plotting the 'Isolation' index
ggplot(ramsar_clean, aes(x = reorder(WetlandName, min_dist_km), y = min_dist_km)) +
  geom_point(color = "#E63946") +
  geom_segment(aes(xend = WetlandName, yend = 0), color = "#CED4DA", alpha = 0.5) +
  coord_flip() +
  labs(title = "Proximity and Ecological Corridors",
       subtitle = "Distance to nearest RAMSAR site (km) - Top 20 most isolated sites",
       x = NULL, y = "Distance to Nearest Neighbor (km)") +
  theme_minimal() +
  scale_x_discrete(breaks = ramsar_clean$WetlandName[seq(1, 96, by = 5)]) # Show every 5th for clarity
```

The "Isolation Index" identifies which sites are "Oases" and which are "Corridors." Sites in **Tamil Nadu and Punjab** show very low "distance to nearest neighbor," indicating they form a functional ecological network. In contrast, the high isolation of sites like **Pala Wetland (Mizoram)** highlights their role as unique, standalone biodiversity refuges that require specialized, independent management strategies.

## Management Diversity & Scale

The dataset names various authorities, from "Retired Academics" to "Forest Departments."

### Does the type of `ManagementAuthority` correlate with the **Total Area** protected? Do institutional authorities manage larger "Mega-Wetlands" while local/academic bodies manage smaller "Niche Sanctuaries"?

```{r}
#| label: Visualizing if large basins are managed differently than smaller bird sanctuaries.
# 1. Simplify Authority Categories
ramsar_auth <- ramsar_clean |>
  mutate(AuthType = case_when(
    str_detect(ManagementAuthority, "Forest|Department|Wildlife") ~ "State Forest Dept",
    str_detect(ManagementAuthority, "University|Academic|Retired") ~ "Academic/Expert",
    TRUE ~ "Other/Local Authority"
  ))

# 2. Visualization
ggplot(ramsar_auth, aes(x = AuthType, y = Area, color = AuthType)) +
  geom_jitter(size = 3, alpha = 0.6, width = 0.2) +
  scale_y_log10(labels = label_comma()) +
  scale_color_manual(values = c("#264653", "#E63946", "#2A9D8F")) +
  labs(title = "Institutional Scale and Oversight",
       subtitle = "Wetland area managed by different authority types",
       x = "Authority Category", y = "Area (Hectares)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(face = "bold"))
```

The jittered scatter plot shows a clear institutional division: **State Forest Departments** manage the widest range of sites, including the "Mega-Wetlands." However, **"Other/Local Authorities"** and **"Academic/Expert"** bodies are increasingly involved in managing smaller, niche sanctuaries. This indicates a decentralization of conservation, where specialized knowledge is being used to protect smaller but ecologically "loud" sites.

### 
