---
title: "Evaluating 40 Years of Wetland Conservation Policy (1981–2025)"
subtitle: "A Spatio-Temporal EDA of India’s RAMSAR Sites"
author: 
  - name: "Shashank Venkatesh"
    email: "shashankv285@gmail.com"
date: last-modified
format:
  html:
    theme: cosmo
    css: styles.css
    toc: true
    toc-depth: 2
    toc-location: left
    code-fold: false
    code-tools: false
    embed-resources: true
    fig-height: 5  
    fig-width: 5 
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
editor_options: 
  chunk_output_type: console

comments:
  giscus:
    repo: "Tailapa/RAMSAR-data-analytics"
    category: "Announcements"  
    mapping: pathname
    reactions-enabled: true
    input-position: top
    theme: light
    language: en
---

::: {#policy_brief .callout style="simple"}
## **Policy Brief: Wetland Governance (1981–2025)**

India’s conservation strategy is driven by the **Amrit Dharohar scheme** and the **National Biodiversity Targets**, aiming to fulfill the **Kunming-Montreal Global Biodiversity Framework (GBF) "30x30" goals.**

Recent judicial interventions, specifically ***M.K. Balakrishnan vs. Union of India***, have transformed wetlands from passive landscapes into protected legal entities, mandating rigorous spatial documentation.

This analysis evaluates the **shift from massive basins to a distributed network of niche habitats**, ensuring regional climate resilience across the subcontinent.
:::

## Introduction

Historically, wetlands were often viewed as **wastelands** or voids in the administrative landscape. However, over the last four decades, a fundamental shift in theperception of these spaces has occurred. Wetlands are now recognized as the **Kidneys** of the landscape—vital hydrological vectors that regulate carbon sequestration, flood mitigation, and biodiversity. The **RAMSAR Convention, signed in 1971**, provided the first global frmework for wetland protection. India’s accession to the convention in 1982 marked the beginning of a structured conservation timeline, transforming local water bodies into nodes of international significance.

Wetlands are no longer peripheral to India’s development agenda; they are central to its climate resilience. As of 2025, India has expanded its network to **96 RAMSAR sites** (totaling over 1.5 million hectares), making it the leader in South Asia for protected wetland density. This exercise is particularly timely due to the recent announcement of **Amrit Dharohar Scheme** by the Government of India to promote the "wise use" of RAMSAR sites through community-led conservation and eco-tourism. Further, under the **Kunming-Montreal Global Biodiversity Framework (GBF)**, India is committed to the "30x30" target—protecting 30% of its land and water by 2030 and this pivots wetlands as the most efficient "units" to meet these targets due to their high carbon density. Additionally, The **State of World's Wetlands** and India's own **National Wetland Decadal Change Atlas** highlights that while site numbers are increasing, they are under severe pressure from urban expansion.

This analysis evaluates the **Spatio-Temporal Efficiency** of 40 years of policy. While the number of sites has exploded recently, our research identifies anchor hubs in conservation—where 45% of the protected area is concentrated in just 3% of the sites. The rationale for this study is further bolstered by a strengthening judicial framework. Landmark rulings, such as the **Supreme Court’s directions in the *M.K. Balakrishnan vs. Union of India* case**, have mandated the mapping and protection of over 200,000 wetlands nationwide. The **Wetlands (Conservation and Management) Rules, 2017**, shifted the administrative authority toward States, making local-level data analysis more critical than ever.

## **Research Objective**

By analyzing the **Size Spectrum** and **Designation Timeline**, this exercise seeks to answer a critical question: *Has India’s policy evolved from protecting "Mega-Basins" to a more resilient, distributed network of "Micro-Habitats"?* Understanding this shift is essential for predicting the future trajectory of India's environmental architecture.

## Data Source

The foundational data for this analysis is sourced from the **Open Government Data (OGD) Platform India** which has a catalog on **Ramsar Sites**. This dataset provides the geometric coordinates, area in hectares, and designation dates for the wetlands, allowing for a precise spatial and temporal analysis.

::: {.callout-note appearance="simple"}
## Data Accessibility

The dataset for this analysis is sourced from the **Open Government Data (OGD) Platform India**.

[Download Dataset](https://www.data.gov.in/catalog/ramsar-sites-details){.btn .btn-outline-primary .btn-sm role="button"}
:::

```{r}
#| label: load the dataset
library(tidyverse)
library(tidymodels)
library(scales) # For formatting legend numbers
library(jsonlite)
library(sf)
library(mapindia)
library(ggrepel) # For non-overlapping labels
library(plotly)
library(tidytext)
library(patchwork) # To combine plots if needed
library(ggwordcloud)
library(viridis)

#library(rnaturalearth) # Error in Indian map with wrong administrative boundaries
#library(rnaturalearthdata) #


# URL of the RAMSAR API
api_url <- "https://indianwetlands.in/wp-content/themes/wetlands/restapi-ramsar.php"

# Fetch and parse the data
# Note: Sometimes government APIs require a 'User-Agent' header to allow access
ramsar_json <- fromJSON(api_url)

# Convert to a data frame
# The API usually returns a nested list; we simplify it here
ramsar_data <- as.data.frame(ramsar_json)

```

## Glimpse of the dataframe

```{r}
#| label: glimpse into the dataframe
#| message: false
#| warning: false
glimpse(ramsar_data)
```

The dataset has **`r nrow(ramsar_data)`** rows and **`r ncol(ramsar_data)`** columns.

## Data Cleaning and Preparation

The dataset on RAMSAR wetlands downloaded from OGD is a json file (not structured), and all the observations are stored in character data type. This is not ideal for Exploratory Data Analysis (EDA), which requires a structured dataset. Hence, we need to clean the data by mutating the observations into appropriate data types that will make it easier to analyse.

```{r}
#| label: Converting chr strings to proper data types
#| message: false
#| warning: false
# Convert character strings to numeric and date types
ramsar_clean <- ramsar_data |>
  mutate(
    Area = as.numeric(Area),
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    Date = as.Date(DesignationDate),
    Year = as.numeric(format(Date, "%Y"))
  ) |>
  arrange(Date) |>
  mutate(cumulative_sites = row_number())
```

## Geographical Distribution of Wetlands in India

```{r}
#| label: Geographical Distribution of wetlands in India_2
#| eval: false
#| include: false
# 2. Get India's high-resolution boundary (sf format)
india_map <- ne_countries(scale = "large", country = "India", returnclass = "sf")

# 3. Convert our RAMSAR dataframe into a spatial object (sf)
# We ensure Latitude/Longitude are numeric first
ramsar_sf <- ramsar_clean |>
  filter(!is.na(Latitude) & !is.na(Longitude)) |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 4. Plotting the Geometric Map
ggplot() +
  # Draw the base map of India with a minimalist geometric style
  geom_sf(data = india_map, fill = "#F8F9FA", color = "#ADB5BD", size = 0.7) +
  
  # Plot wetlands as geometric bubbles
  # Size is proportional to the Area; color is our signature Teal
  geom_sf(data = ramsar_sf, 
          aes(size = Area), 
          color = "#2A9D8F", 
          alpha = 0.6) +
  
  # Add labels only for the top 5 largest sites to keep it "geometric and clean"
  geom_text_repel(data = ramsar_sf |> slice_max(Area, n = 5),
                  aes(label = WetlandName, geometry = geometry),
                  stat = "sf_coordinates",
                  size = 3, 
                  fontface = "bold",
                  family = "sans",
                  color = "#264653",
                  box.padding = 0.5) +
  
  # Formatting the scales and legend
  scale_size_continuous(range = c(1, 12), 
                        labels = label_comma(),
                        name = "Area (Hectares)") +
  
  # Apply a minimalist void theme
  theme_void() +
  labs(
    title = "FIG 1: Dsitribution of India's Ramsar Sites",
    subtitle = "Analysis of 96 internationally significant wetlands as of Dec 2025",
    caption = "Data Source: Open Government Data (OGD) Platform India | Projection: WGS84"
  ) +
  
  # Styling the typography
  theme(
    plot.title = element_text(family = "sans", face = "bold", size = 18, color = "#264653", hjust = 0.5),
    plot.subtitle = element_text(family = "sans", size = 11, color = "#666666", hjust = 0.5, margin = margin(b = 20)),
    legend.position = c(0.85, 0.2),
    plot.background = element_rect(fill = "white", color = NA)
  )

```

```{r}
#| label: Geographical Distribution of wetlands in India
#| fig-height: 7  # Increases the vertical height of the output
#| fig-width: 7   # Increases the horizontal width
#| dpi: 300        # High resolution for geometric clarity

# 1. Retrieve the official state boundaries as an sf object
# This function is specifically designed to provide the official Indian map
india_official <- map_india(regions = "states")

#Convert our RAMSAR dataframe into a spatial object (sf)
# We ensure Latitude/Longitude are numeric first
ramsar_sf <- ramsar_clean |>
  filter(!is.na(Latitude) & !is.na(Longitude)) |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 2. Plotting the correct geometry
ggplot() +
  # Draw the base map with the official boundaries
  geom_sf(data = india_official, fill = "#F8F9FA", color = "#ADB5BD", linewidth = 0.3,size = 0.4) +
  
  # Layer the RAMSAR sites as geometric nodes
  geom_sf(data = ramsar_sf, 
          aes(size = Area), 
          color = "#2A9D8F", 
          alpha = 0.6) +
  scale_size_continuous(range = c(2, 12), name = "Area (sq. km)") +
  coord_sf(expand = FALSE) +
  
  theme_void() +
  labs(
    title = "Geospatial Distribution of RAMSAR Sites",
    subtitle = "40 Years of Policy: Analyzing 96 RAMSAR sites",
    caption = "Data Source: mapindia package & OGD India"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#264653", hjust = 0.5, vjust = 4), plot.subtitle = element_text(size = 11, color = "#666666", hjust = 0.5, vjust = 4), 
    legend.position = "bottom",
    plot.margin = margin(10, 10, 10, 10) # Adds internal breathing room
  )
```

The above map illustrates that India's RAMSAR sites are not uniformly distributed but are clustered. We can see a concentration of sites in the **Indo-Gangetic Plain** and a significant concentration along the **Southern Coast (Tamil Nadu)**. The nodal points also indicate that the largest wetlands (by area) are mostly located in the North and East, while the South is characterized by a cluster of smaller sites.

## The Acceleration of Conservation: A Temporal Growth Analysis

### How has the rate of RAMSAR designations in India changed over the last three decades, particularly since the 2020s?

to pinpoint the "inflection point" where India’s wetland conservation policy shifted into high gear.

```{r}
#| label: temporal growth analysis
#| message: false
#| warning: false
ggplot(ramsar_clean, aes(x = Date, y = cumulative_sites)) +
  #geom_area(fill = "#2A9D8F", alpha = 0.2) +
  geom_line(color = "#2A9D8F", size = 0.5) +
  geom_point(data = filter(ramsar_clean, Year >= 2020), 
             color = "#E63946", size = 1) +
  labs(
    title = "The Exponential Rise of Indian RAMSAR Sites",
    subtitle = "Cumulative designations from 1981 to December 2025",
    x = "Year of Designation", y = "Total Number of Sites"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 16, color = "#264653"))
```

It is apparent from the above plot that a **Policy Inflection Point may have occurred around 2020**. While designations were sporadic from 1981 to 2010, the curve steepens sharply in the last five years with more than half (59) of India's current Ramsar sites designated since 2020, signaling an aggressive shift in national environmental policy and a heightened commitment to international conservation standards.

```{r}
#| label: temporal analysis of designation 

# 1. Categorize the sites based on the 2020 inflection point
ramsar_phases <- ramsar_clean |>
  mutate(Phase = ifelse(Year < 2020, "Before 2020", "Since 2020")) |>
  count(Phase) |>
  mutate(Percentage = n / sum(n) * 100)

# 2. Print the summary table to console
#print(ramsar_phases)

# 3. Geometric Vector Visualization
ggplot(ramsar_phases, aes(x = Phase, y = n, fill = Phase)) +
  # Using a fixed width for a clean geometric look
  geom_col(width = 0.6, color = "#264653", size = 1) +
  # Adding count labels directly on top of the bars
  geom_text(aes(label = n), vjust = -0.5, size = 6, fontface = "bold", color = "#264653") + 
  ylim(0, 70) +
  # Applying the Teal and Coral palette
  scale_fill_manual(values = c("Before 2020" = "#2A9D8F", "Since 2020" = "#E63946")) +
  labs(
    title = "PHASE ANALYSIS: THE 2020 CONSERVATION PIVOT",
    subtitle = "Comparing decades of legacy designations vs. recent geometric growth",
    x = NULL, 
    y = "Number of Designated Sites"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16, color = "#264653"),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 12, face = "bold", color = "#264653")
  )
```

As the plot confirms, before 2020 (The Legacy Era), we only had 37 RAMSAR sites. This represents nearly 40 years of gradual identification and designation since India joined the convention in 1982. However, post-2020 (The Modern Era), we can see that 59 additional sites were designated as RAMSAR sites, which forms **61% of India’s total RAMSAR sites were declared in just the last 5 years.** This sudden surge indicates a shift in environmental priority, where the administrative speed of protecting wetlands has tripled compared to the previous four decades.

## State-wise Distribution of Wetlands

To understand thearchitecture of India’s Ramsar conservation, we must look beyond numbers to examine the state-wise distribution. We need to understand here that if the distribution is concentrated, the national conservation strategy would rely on a few power clusters, whereas in a dispersed distribution, the responsibility to manage is also distributed, creating a more economic regulatory framework.

The designation of a Ramsar site is a reflection of both ecological wealth and administrative proactivity. Mapping the site count across states would enable the identification of **clustered corridors** and the **sparse zones** of the Ramsar network in Indian context. Therefore, we need to determine if a small number of states hold the majority of India's RAMSAR designations, or is it evenly distributed.

```{r}
#| label: State-wise Distribution

ramsar_clean |>
  count(StateName) |>
  mutate(Highlight = ifelse(n >= 10, "Top Tier", "Other")) |>
  ggplot(aes(x = n, y = reorder(StateName, n), fill = Highlight)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("Top Tier" = "#E63946", "Other" = "#2A9D8F")) +
  labs(
    title = "Regional Distribution of Wetlands",
    subtitle = "Tamil Nadu leads the geometric distribution with 20 sites",
    x = "Number of Designated Sites", y = NULL, fill = "Status"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 9, face = "bold"))
```

```{r}
#| label: states with highest wetland sites
#| echo: false

top_states <- ramsar_clean |>
  group_by(StateName) |>
  summarise(TotalSites = n()) |>
  arrange(desc(TotalSites))

top_two <- percent(sum(top_states$TotalSites[1:2])/sum(top_states$TotalSites[1:nrow(top_states)]), accuracy = 0.1)
top_five <- percent(sum(top_states$TotalSites[1:5])/sum(top_states$TotalSites[1:nrow(top_states)]), accuracy = 0.1)


niche_states_data <- ramsar_clean |>
  group_by(StateName) |>
  summarise(SiteCount = n()) |>
  filter(SiteCount %in% c(1, 2)) 

summary_counts <- niche_states_data |>
  group_by(SiteCount) |>
  summarise(NumberOfStates = n())

total_states = nrow(top_states)

one_or_two_states = sum(summary_counts$NumberOfStates[1:2])
one_or_two_states_p = percent(one_or_two_states/total_states, accuracy = 0.1)
```

The bar plot confirms a disparity in wetland distribution. **Tamil Nadu** leads the country with **`r top_states$TotalSites[1]` sites**, followed by Uttar Pradesh (**`r top_states$TotalSites[2]`**). These top two states have **`r top_two`** of the total number of RAMSAR sites in India. Further, the share of the top 5 states in this metric is an overwhelming —**`r top_five`**. Conversely, **`r one_or_two_states_p` of states (`r one_or_two_states`)** have only 1or 2 sites, suggesting that conservation success is currently driven by specific state-level initiatives rather than a perfectly uniform national spread.

## The Size Spectrum: Are we protecting large basins or small habitats?

As established in Fig:1, it is clear that the area represented by Ramsar sites is not uniform throughout the country. This aligns with India's diverse geographical terrain where different habitats are nurtured by diferent types of wetlands, but all of them are necessary to sustain their ecosystem.

For instance, larger sites like the **Sundarbans (West Bengal)** or **Vembanad-Kol (Kerala)** act as macro-anchors which are essential for regional climate regulation, carbon sequestration, and protecting inland populations from cyclonic surges. Conversly, **medium sized sites** such as **Keoladeo National Park (Rajasthan)** function as nodal hubs, supporting migratory flyways. Whereas, **smaller sites** like the **Renuka Lake (Himachal Pradesh)**, act as refugee/niche habitats creating unique micro-climates for endemic species that would be lost in a larger, more competitive basin.

The key question therefore, is to ask whether there is a correlation in the distribution of wetland sizes (Area), and are there mega-wetlands that skew the national average?

```{r}
#| label: Large basins v/s small habitats
#| eval: false
#| include: false
ggplot(ramsar_clean, aes(x = Area)) +
  geom_histogram(fill = "#264653", color = "white", bins = 20) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Distribution of Wetland Area (Log Scale)",
    subtitle = "Analysis of size variance from small reserves to massive basins",
    x = "Area in Hectares (Log10)", y = "Frequency of Sites"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())


```

```{r}
#| label: calculate-size-spectrum
#| echo: false
#| fig-height: 6
#| fig-width: 9

# 1. Define the geometric brackets using 'cut'
size_summary <- ramsar_clean |>
  mutate(
    Size_Bracket = cut(Area, 
                       breaks = c(0, 10, 100, 1000, 10000, 100000, Inf),
                       labels = c("< 10", "10–100", "100–1,000", "1,000–10,000", "10k–100k", "> 100k"),
                       right = FALSE)
  ) |>
  group_by(Size_Bracket) |>
  summarise(Count = n()) |>
  mutate(Percentage = round((Count / sum(Count)) * 100, 1))



ggplot(size_summary, aes(x = Size_Bracket, y = Count, fill = Size_Bracket)) +
  # Draw clean geometric bars
  geom_bar(stat = "identity", width = 0.7, color = "white", linewidth = 0.5) +
  
  # Add text labels on top of bars
  geom_text(aes(label = Count), vjust = -0.5, size = 5, fontface = "bold", color = "#264653") +
  
  # Professional Color Palette (Teal to Coral)
  scale_fill_manual(values = c("#2A9D8F", "#8AB17D", "#E9C46A", "#F4A261", "#E76F51", "#E63946")) +
  
  ylim(0,40) +
  
  theme_minimal() +
  labs(
    title = "THE SIZE SPECTRUM",
    subtitle = "Analysis of 96 RAMSAR sites by Area Bracket (Hectares)",
    x = "Area Bracket (Ha)", y = "Number of Sites",
    fill = "Size Category"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16, color = "#264653", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text = element_text(size = 11, color = "#264653"),
    panel.grid.major.x = element_blank()
  )
```

The bar plot shows a shift in India's wetland protection architrcture from broad landscape protection to targeted biodiversity conservation. **Most sites (38.5%** of total number of Ramsar sites) fall within the **1,000 to 10,000-hectare range.** However, the presence of sites on the extreme right of the plot indicates mega-wetlands that serve as critical national ecological anchors (discussed in the next section), while the smaller peaks represent specialized bird sanctuaries or niche habitats.

**The "Mid-Range" Dominance (The 100–1,000 Hectare)**

**38.5%** of all Indian RAMSAR sites fall into the **100–1,000 hectare range**. This suggests that India’s conservation framework is optimized for medium-sized water bodies. These are often large enough to support significant migratory bird populations but small enough to be managed under specific state-level administrative boundaries.

**The Mega-Wetlands are rare**

Only **3.1%** (3 sites) exceed 100,000 hectares. In a 40-year policy evaluation, we can consider them as legacy sites. For example, wetland ecosystems like the Sundarbans. Their low count isn't a failure of policy, but rather a reflection of the geographic reality: there are very few ecosystems of this magnitude in the country. They act as the **National Anchors** of the network.

**The Rise of Small Habitats (10–100 Hectares)**

With **15.6%** of sites being relatively small (\< 100 ha), there is clear evidence of **Targeted Biodiversity Conservation**. In recent years (2020–2025), policy has shifted toward identifying "Micro-Nodes," sites that may be small in area but are critical for migratory paths like the Central Asian Flyway. For example, **Kazhuveli Bird Sanctuary in Tamilnadu.**

*The conservation architecture of India's RAMSAR sites indicate a shift from broad 'Landscape Protection' to 'Targeted Biodiversity Conservation.' While the early decades focused on securing massive basins (\>100k ha), the recent designations are defined by the proliferation of mid-sized (100–1,000 ha) and micro-sites (\<100 ha), emphasizing a more granular and inclusive approach to wetland management.*

## 3 vs 93: The Anchor Sites

As previously discussed, just 3 sites (3.1% of total number of sites) hold a substantial area, each exceeding 100,000 hectares. We need to examine this further to analyse how these 3 large sites differ compared to other smaller sites on the Ramsar network.

```{r}
#| label: area-concentration-analysis
#| echo: false

# 1. Sort by area and calculate cumulative share
area_dist <- ramsar_clean |>
  arrange(desc(Area)) |>
  mutate(
    cum_area = cumsum(Area),
    total_area = sum(Area),
    cum_perc = (cum_area / total_area) * 100,
    rank = row_number()
  )

# 2. Extract specific stats for the Top 3 vs the Rest
top_3_stats <- area_dist |> 
  filter(rank <= 3) |> 
  summarise(
    Count = n(),
    Share_of_Total_Area = max(cum_perc)
  )

# 1. Prepare the comparison data
comparison_df <- ramsar_clean |>
  arrange(desc(Area)) |>
  mutate(Category = ifelse(row_number() <= 3, "3 sites", "93 sites")) |>
  group_by(Category) |>
  summarise(Total_Area = sum(Area)) |>
  mutate(
    percentage = Total_Area / sum(Total_Area),
    label = paste0(Category, "\n", comma(round(Total_Area)), " ha\n(", round(percentage * 100, 1), "%)")
  )

# 2. Plotting the Geometric Donut
ggplot(comparison_df, aes(x = 2, y = percentage, fill = Category)) +
  # The "Ring" vector
  geom_bar(stat = "identity", width = 1, color = "white", linewidth = 2) +
  coord_polar(theta = "y") +
  
  # Create the "Donut" hole (Inner Radius)
  xlim(0.5, 2.5) + 
  
  # Theme Colors (Coral for Anchors, Teal for Nodes)
  scale_fill_manual(values = c("3 sites" = "#E63946", "93 sites" = "#2A9D8F")) +
  
  # Styling and Labels
  theme_void() +
  labs(
    title = "THE ANCHOR WETLANDS",
    subtitle = "Areal breakdown of India's 40-year RAMSAR footprint",
    caption = "3.1% of sites vs 96.9% of sites"
  ) +
  # Overlaying the Data Labels
  geom_text(aes(label = label), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            fontface = "bold", 
            size = 4) +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = "#264653", hjust = 0.5),
    plot.subtitle = element_text(size = 12, color = "#666666", hjust = 0.5, margin = margin(b = 20)),
    plot.caption = element_text(size = 10, hjust = 0.5),
    legend.position = "none" # Labels are inside the chart
  )
```

The most crucial finding of this EDA is the extreme concentration of protected acreage in just three locations - `r area_dist$WetlandName[1]`, `r area_dist$WetlandName[2]` and `r area_dist$WetlandName[3]` . These **3 sites** (3.1% of the total Ramsar sites) command nearly **half (44.9%)** of the entire national RAMSAR footprint. With **690,750 hectares** concentrated in these hubs, they serve as the primary Carbon sinks and hydrological regulators for the country. This anchors them as the most crucial hubs in India's Ramsar sites conservation framework and hence these sites warrant special and focused attention. The remaining **93 sites** represent **847,207 hectares**, averaging only about 9,100 hectares per site. While these sites are individually smaller, they act as spokes of the wetland protection network since they provide vital connectivity across different climatic zones, bird flyways, and state borders.

*Overall, while the Hubs provide the scale and act as national ecological anchors, the Spokes provide the resilience by ensuring that if one region faces a drought or ecological crisis, other nodes in the network can still support migratory species. This **dual-track strategy** of protecting massive hubs while expanding a dense network of small spokes is the highlight of India's modern conservation framework.*

## Relationship between Latitude & Area of Wetlands

The objective here is to investigate if northern wetlands tend to be larger or smaller than wetlands in the peninsula and to analyse if there exists a correlation between geographical latitude and the size of designated wetlands. We can do this by comparing the area of wetlands against their coordinates. This EDA reveals a surprising **inversion** in India’s wetland conservation paradigm. Despite the North having a network of rivers like the Indus-Ganga floodplains, the **Peninsular South** actually holds the overwhelming majority of India's protected wetland mass.

```{r}
#| label: north-south-split
#| echo: false

# 1. Define the Divide (Tropic of Cancer ~23.5)
divide_lat <- 23.5

comparison_ns <- ramsar_clean |>
  mutate(Region = ifelse(Latitude >= divide_lat, "North (Above 23.5°N)", "South (Below 23.5°N)")) |>
  group_by(Region) |>
  summarise(
    Site_Count = n(),
    Total_Area = sum(Area),
    Average_Area = mean(Area),
    Median_Area = median(Area)
  ) |>
  mutate(Area_Percentage = round((Total_Area / sum(Total_Area)) * 100, 1))

ggplot(ramsar_clean, aes(x = Latitude, y = Area)) +
  # 1. The Data Points (Teal Nodes)
  geom_point(aes(size = Area), color = "#2A9D8F", alpha = 0.5) +
  
  # 2. The Statistical Trend (Coral Vector)
  geom_smooth(method = "lm", color = "#E63946", se = FALSE, linetype = "solid", linewidth = 1) +
  
  # 3. The Geometric Equator (Tropic of Cancer)
  geom_vline(xintercept = 23.5, linetype = "dashed", color = "#264653", linewidth = 0.8) +
  
  # 4. Adding a label for the Threshold
  annotate("text", x = 24, y = 130, label = "Tropic of Cancer (23.5°N)", 
           angle = 90, color = "#264653", fontface = "italic", size = 3.5) +
  
  # Scaling and Design
  scale_y_log10(labels = comma) +
  labs(
    title = "Relationship between Latitude and Wetland Size",
    subtitle = "Isolating Northern Basins from Peninsular Nodes",
    x = "Latitude (Degrees North)", y = "Area in Hectares (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = "#264653", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    panel.grid.minor = element_blank()
  ) +
  guides(size = "none")
```

**Numerical Parity**

Geographically, there is almost perfect balance in terms of distribution with the North (above 23.5°N) having 46 sites and the South (below 23.5°N) having 50 sites. This points to a symmetric designation of sites across India. We can hence conclude that our "Tropic of Cancer (ToC)" line acts as an almost perfect axis of symmetry for the number of designations.

**Area Asymmetry**

Surprisingly though, **79.3% of total wetland area is concentrated** in the South. Despite having a similar number of sites, the **South** holds **1,219,308 hectares**, nearly **4x the total area** of the North (**318,648 hectares**). The "Average Area" in the South ($24,386$ ha) is significantly higher than in the North (6,927 ha), driven by Anchor Hubs like the Sundarbans and Chilika Lake.

**Trend Line**

As we can infer from the plot, there is a slight increase in size as we go North. However, the **high variance** (the distance of dots from the line) is the **real story.** This **outlier effect can be** noticed from the massive nodes just to the left of the Tropic of Cancer. This exerts more gravitational pull on the dataset than 20 smaller sites in the far North combined.

**Median vs. Mean**

The fact that the Medians are so close (624 ha for north and 761 ha for south) while the Means are so far apart (6,927 vs. 24,386) proves that Indian wetland conservation architecture is anchored in a few massive sites.

*Overall, while Northern India provides a consistent network of 46 sites, Peninsular India provides a relatively similar network of 50 sites but represents nearly 80% of the total protected area. Our EDA analysis reveals that conservation below ToC is defined by **massive basins**, while above ToC, it is defined by **niche clusters**.*

## District-Level Granularity within Leading States

Within the states with the most RAMSAR sites, are the wetlands clustered in specific districts or spread across the state?

Identifies "ecological hubs" at a regional level, which is useful for management authority analysis.

```{r}
#| label: Identifying ecological hubs
ramsar_clean |>
  filter(StateName == "Tamil Nadu") |>
  count(districtname) |>
  ggplot(aes(x = n, y = reorder(districtname, n))) +
  geom_segment(aes(x = 0, xend = n, yend = districtname), color = "#2A9D8F") +
  geom_point(size = 4, color = "#E63946") +
  labs(
    title = "Local Concentration: Tamil Nadu",
    subtitle = "District-wise breakdown of sites in India's leading RAMSAR state",
    x = "Number of Sites", y = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

The scatter plot with the regression line shows a fascinating geospatial trend: **Northern wetlands tend to be larger in area.** As latitude increases, we see more "Mega-Wetlands." This is likely due to the vast floodplains of the Himalayan river systems. In contrast, southern latitudes (closer to the equator) feature a higher number of smaller, fragmented coastal and tank ecosystems.

## Natural vs. Human-Made Wetlands: Is there an Anthropogenic shift?

Recent additions made in 2025, like **Siliserh Lake** (Rajasthan) and **Kopra Reservoir** (Chhattisgarh), are artificial ecosystems. This analysis reveals if India is increasingly relying on man-made reservoirs for international conservation status and determines if India’s conservation strategy is shifting toward protecting infrastructure-based ecosystems. Our goal here is to investigate what percentage of India's RAMSAR network is comprised of human-made "Artificial" wetlands, and to evaluate how their average size compare to natural basins.

```{r}
#| label: Artificial wetlands

# 1. Classification Logic
ramsar_cat <- ramsar_clean |>
  mutate(Type = case_when(
    str_detect(WetlandName, "Reservoir|Dam|Barrage|Tank|Canal|Reserve") ~ "Human-Made / Managed",
    TRUE ~ "Natural / Semi-Natural"
  ))

summary_type <- ramsar_cat |>
  group_by(Type) |>
  summarise(
    Count = n(),
    Total_Area = sum(Area),
    Mean_Area = mean(Area),
    Median_Area = median(Area)
  ) |>
  mutate(Area_Share_Pct = round((Total_Area / sum(Total_Area)) * 100, 1))


# 2. Visualization
ggplot(ramsar_cat, aes(x = Type, y = Area, fill = Type)) +
  geom_violin(alpha = 0.3, color = NA) +
  geom_boxplot(width = 0.1, color = "#264653", outlier.shape = NA) +
  geom_jitter(aes(color = Type), alpha = 0.5, width = 0.15) +
  scale_y_log10(labels = label_comma()) +
  scale_fill_manual(values = c("Human-Made / Managed" = "#E63946", "Natural / Semi-Natural" = "#2A9D8F")) +
  scale_color_manual(values = c("Human-Made / Managed" = "#E63946", "Natural / Semi-Natural" = "#2A9D8F")) +
  labs(title = "Natural vs. Managed Ecosystems",
       subtitle = "Comparison of total area distribution (Hectares)",
       x = NULL, y = "Area (Hectares, Log Scale)") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
    )
```

The analysis makes it clear that India’s Ramsar portfolio is overwhelmingly focused on **Natural/Semi-Natural** systems with a commanding majority of **83 sites**, while **Human-made/Managed sites** (tanks and reservoirs) represent just **13 sites**. We can therefore say that India's conservation architecture is built on an organic foundation with human-made sites representing a supplementary layer of conservation, rather than being the primary intervention strategy.

Moreover, one of the critical discoveries stemming from this analysis is that the **median area** of a typical human-made Ramsar site (1,675 ha) is **over three times larger** than a natural one (526 ha). Nonetheless, the **natural sites still represent 88.6% of the total area** compared to just 11.4% shared by human-made sites.

It should also be noted here that the human-made reservoirs and barrages are now statistically comparable in importance to natural lakes, since India is leveraging its irrigation infrastructure to provide internationally recognized habitats for migratory species. Examples include **Pong dam in Himachal Pradesh** and **Beas Conservation Reserve in Punjab.**

## Semantic Analysis of Uniqueness

A good approach to understand the uniqueness of India's RAMSAR sites is by performing **Natural Language Processing (NLP)** on the **qualitative data (categorical** data) contained in our dataset. This can be done by extracting and identifying specific keywords from the **observation "Uniqueness"** of the dataset that justify why these 96 sites were recognized by awarding them with international status.

The key question here is to evaluate the top three Ecological Drivers (e.g., Birds, Flood Control) that qualify an Indian site for RAMSAR status. This can be done by extracting the keywords of Uniqueness Observation to see what makes a wetland internationally significant in the Indian context.

```{r}
#| label: Geometric Word Cloud
#| echo: false

# 1. Update the list with strictly lowercase terms
exclude_words <- c("wetland", "ramsar", "india", "sites", "lake", "water", "ground")

# 2. Text Mining
uniqueness_words <- ramsar_clean |>
  unnest_tokens(word, Uniqueness) |>
  anti_join(stop_words) |>
  filter(!word %in% exclude_words) |>
  count(word, sort = TRUE) |>
  head(15)

# 1. Generate the Geometric Word Cloud
ggplot(uniqueness_words, aes(label = word, size = n, color = n)) +
  # Using the 'area' shape for a more geometric, structured distribution
  geom_text_wordcloud_area(area_corr = TRUE, shape = "square", rm_outside = TRUE) +
  
  # Scaling the text size geometrically
  scale_size_area(max_size = 50) +
  
  # Applying Teal and Coral color palette
  scale_color_gradient(low = "#2A9D8F", high = "#E63946") +
  
  # Minimalist formatting
  theme_void() +
  theme(
    plot.background = element_rect(color = NA)
  )
```

```{r}
#| label: analyse the uniqueness

# 2. Geometric Bar Plot
ggplot(uniqueness_words, aes(x = n, y = reorder(word, n))) +
  geom_col(fill = "#264653") +
  geom_text(aes(label = n), hjust = -0.2, size = 3, color = "#264653") +
  labs(title = "The Vocabulary of Conservation",
       subtitle = "Top recurring terms in site uniqueness descriptions",
       x = "Frequency of Mention", y = NULL) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

The high frequency of terms like **"bird/birds/waterbirds" (appears 14 times),** **"sanctuary/reservoir" (appears 9 times)**, and **"migratory" (appears 4 times)**, confirms that India’s RAMSAR sites are primarily designated because of their role as bird sanctuaries in the Central Asian Flyway. This also signals that a large portion of India’s protected wetlands are irrigation reservoirs and designated sanctuaries, integrated into international conservation frameworks.

Further, the appearance of words like **"freshwater" (5 times)** and **"endangered" (3 times)** indicates that the often used logic behind India's conservation framework of wetlands is threat-response, where the attention is focused on inland freshwater ecosystems (not just coastal) and their role in preventing the extinction of specific species.

## The Seasonality of Global Recognition

As previously discussed, at the outset, there were **only 37 sites before 2020**. This **changed drastically** in the **last five years** with the announcement of **59 additional sites**, which often involved **bundled designations** in one go. This makes one curious to inquire if there is a **"policy window"** where sites are more likely to be designated around specific international events (e.g., World Wetlands Day or Independence Day). The answer to this would show **if political timelines influence environmental milestones**. We can understand this by employing a **radial plot** to see if designations cluster around specific months (Policy Momentum).

```{r}
#| label: seasonality-analysis
ramsar_months <- ramsar_clean |>
  mutate(Month = month(Date, label = TRUE)) |>
  count(Month) |>
  arrange(desc(n))

ggplot(ramsar_months, aes(x = Month, y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_polar() +
  scale_fill_gradient(low = "#8ECAE6", high = "#219EBC") +
  labs(title = "The Annual Cycle of Designations",
       subtitle = "Frequency of RAMSAR announcements by month",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

As can be inferred from the radial plot, the announcement of RAMSAR sites are not distributed evenly throughout the year. Rather, we can observe a nesting in a few months, like **`r ramsar_months$Month[1]`,** when **`r ramsar_months$n[1]`** sites were declared, and in **`r ramsar_months$Month[2]`** when **`r ramsar_months$n[2]`**, and in **`r ramsar_months$Month[3]`,** when again, **`r ramsar_months$n[3]`** sites were declared as RAMSAR sites. This aligns with prominent annual events such as **World Wetlands Day on February** **2nd** or **Independence day on August 15th**. For instance, on India's 75th Independence Day, Environment Minister Sri Bhupendra Yadav declared on social media pllatform 'X' that he was, "Elated to inform that 11 more Indian wetlands have got RAMSAR recognition. This takes our tally to 75 sites." \[1\] This shows that there is a window of announcement, often tied to global environmental milestones and diplomatic postures.

## Fragmented vs. Clustered Landscapes

Considering the spatial distribution of Ramsar sites, one might argue that there is no clear pattern as to whether they are getting clustered or isolated. Hence, the question we pose here is to investigate the pattern of spatial distribution of Ramsar sites to identify Hotspots (like the Tamil Nadu belt) versus Isolated sites (like Pala Wetland in Mizoram). This can be done by calculating how far each site is from its closest neighbor. The findings are interesting.

```{r}
#| label: Nearest Neighbor Identification
#| fig-height: 10
#| fig-width: 8

# --- 1: SPATIAL GEOMETRY CALCULATIONS ---
# 1.1 Extract coordinates
coords <- st_coordinates(ramsar_sf)

# 1.2 Generate the Distance Matrix (Euclidean distance in meters)
dist_matrix <- st_distance(ramsar_sf)

# 1.3 Mask the diagonal (Self-distance) to find ACTUAL neighbors
diag(dist_matrix) <- Inf 

# 1.4 Calculate min distance and find neighbor indices
# Convert meters to kilometers for research standard units
ramsar_clean$min_dist_km <- apply(dist_matrix, 1, min) / 1000
nearest_neighbor_index <- apply(dist_matrix, 1, which.min)

# 1.5 Map Neighbor names back to the dataset
ramsar_clean <- ramsar_clean |>
  mutate(
    Nearest_Neighbor = WetlandName[nearest_neighbor_index]
  )


# --- STEP 2: DATA PREPARATION FOR GEOMETRIC PLOT ---
# We focus on the Top 20 most isolated sites to maintain design clarity
isolated_top_20 <- ramsar_clean |>
  slice_max(min_dist_km, n = 20) |>
  mutate(
    # Truncate long names for the Y-axis (Typography Management)
    ShortName = ifelse(nchar(WetlandName) > 30, 
                       paste0(substr(WetlandName, 1, 27), "..."), 
                       WetlandName),
    # Create a label for the plot that includes the neighbor name
    Label_Full = paste0(ShortName, " (Near: ", Nearest_Neighbor, ")")
  )


# --- STEP 3: GEOMETRIC VECTOR VISUALIZATION ---
ggplot(isolated_top_20, aes(x = reorder(ShortName, min_dist_km), y = min_dist_km)) +
  
  # The "Stem" (Geometric Vector)
  geom_segment(aes(xend = reorder(ShortName, min_dist_km), yend = 0), 
               color = "#CED4DA", 
               linewidth = 1) +
  
  # The "Node" (Data Point) - Color intensity mapped to distance
  geom_point(aes(color = min_dist_km), size = 4) +
  
  # Professional Color Scale: Teal (Connected) to Coral (Isolated)
  scale_color_gradient(low = "#2A9D8F", high = "#E63946") +
  
  # Layout Flip for Readability
  coord_flip() +
  
  # Academic Labels
  labs(
    title = "ECOLOGICAL ISOLATION: TOP 20 VECTORS",
    subtitle = "Quantifying fragmentation: Distance (km) to the nearest neighbor",
    x = NULL, 
    y = "Minimum Separation Distance (km)",
    color = "Isolation (km)"
  ) +
  
  # Geometric Theme Styling
  theme_minimal() +
  theme(
    text = element_text(family = "sans", color = "#264653"),
    plot.title = element_text(face = "bold", size = 16),
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank(), # Clean background
    legend.position = "bottom"
  )

# --- STEP 4: OUTPUT DATA TABLE ---
# This creates a summary table for your Quarto site
isolation_summary <- ramsar_clean |>
  select(WetlandName, Nearest_Neighbor, min_dist_km) |>
  arrange(desc(min_dist_km))

```

Sites like **Kolleru Lake (423 km** away from the nearest wetland), **Deepor Beel (268 km)**, and **Loktak Lake (268 km)** are what experts might call **"Ecological Islands."** These sites are geographically lonely considering the nearest RAMSAR site. Hence, from a policy perspective, this necessitates localized, high-intensity management because they cannot rely on neighboring wetlands.

At the other end of the spectrum, we have sites like **Chitrangudi and Kanjirankulam (2.07 km)** or **Nagi and Nakti (5.34 km)**. These are not just individual sites; they are **clustered complexes**. Geometrically, these nodes are so close they likely share the same migratory bird populations and water tables. This suggests that recent policy (post-2020) has shifted toward protecting "clusters," effectively creating larger safety zones rather than just isolated spots.

## Governing Diversity

The dataset names various authorities, from "Retired Academics" to "Forest Departments." One critical question to ask is whether there is a correlation between the type of wetland, its area and management authority. The answer to this question would reveal if institutional authorities manage mega-wetlands and whether local/academic bodies manage smaller niche sanctuaries.

```{r}
#| label: Visualizing if large basins are managed differently than smaller bird sanctuaries.
#| include: false
#| eval: false

# 1. Simplify Authority Categories
ramsar_auth <- ramsar_clean |>
  mutate(AuthType = case_when(
    str_detect(ManagementAuthority, "Forest|Department|Wildlife") ~ "State Forest Dept",
    str_detect(ManagementAuthority, "University|Academic|Retired") ~ "Academic/Expert",
    TRUE ~ "Local Authority"
  ))

governance_summary <- ramsar_auth |>
  group_by(AuthType) |>
  summarise(
    Count = n(),
    Median_Area = median(Area),
    Total_Area = sum(Area),
    .groups = "drop"
  ) |>
  mutate(Area_Contribution = round((Total_Area / sum(Total_Area)) * 100, 2))

print(governance_summary)

# 2. Visualization
ggplot(ramsar_auth, aes(x = AuthType, y = Area, color = AuthType)) +
  geom_jitter(size = 3, alpha = 0.6, width = 0.2) +
  scale_y_log10(labels = label_comma()) +
  scale_color_manual(values = c("#264653", "#E63946", "#2A9D8F")) +
  labs(title = "Institutional Scale and Oversight",
       subtitle = "Wetland area managed by different authority types",
       x = "Authority Category", y = "Area (Hectares)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(face = "bold"))
```

```{r}
#| label: governance-heatmap
#| fig-height: 6
#| fig-width: 7
#| echo: false

# 1. Combine all classification vectors
ramsar_matrix <- ramsar_clean |>
  mutate(
    Type = case_when(
      str_detect(WetlandName, "Reservoir|Dam|Barrage|Tank|Canal|Reserve") ~ "Human-Made",
      TRUE ~ "Natural"
    ),
    AuthType = case_when(
      str_detect(ManagementAuthority, "Forest|Department|Wildlife") ~ "State Forest Dept",
      str_detect(ManagementAuthority, "University|Academic|Retired") ~ "Academic/Expert",
      TRUE ~ "Other/Local Authority"
    )
  )

# 2. Calculate the Geometric Correlation Table
governance_summary <- ramsar_matrix |>
  group_by(AuthType, Type) |>
  summarise(
    Count = n(),
    Median_Area = median(Area),
    Total_Area = sum(Area),
    .groups = "drop"
  ) |>
  mutate(Area_Contribution = round((Total_Area / sum(Total_Area)) * 100, 2))

# 1. Prepare the Grid Data
heatmap_data <- ramsar_matrix |>
  group_by(AuthType, Type) |>
  summarise(Total_Area = sum(Area), .groups = "drop")

# 2. Plotting the Geometric Matrix
ggplot(heatmap_data, aes(x = Type, y = AuthType, fill = Total_Area)) +
  geom_tile(color = "white", linewidth = 0.8) + # Creates the grid border
  scale_fill_gradient(
    trans = "log10", 
    low = "#e9ecef", high = "#264653", # From neutral to "Massive" Dark Teal
    labels = comma
  ) +
  geom_text(aes(label = comma(round(Total_Area))), color = "white", fontface = "bold", size = 6) +
  labs(
    title = "Governance-Area-Origin Matrix",
    subtitle = "Total Area (Hectares) across Institutional Actors",
    x = "Ecological Origin", y = "Management Authority",
    fill = "Total Area (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

The analysis indicates the general trend mapping key governing institutions managing Ramsar sites based on their type and their area.

**Natural Sites**

For instance, With **54 sites** and managing **60% of the total area (**923,172 ha**)**, **State Forest Departments** emerge as the dominant governing authority for natural Ramsar sites. On the other hand, the small median of 488 ha for this group shows that the State Departments manages a large tail of small Natural Ramsar sites, anchored by a few larger sites.

Unlike the state departments, **Other/Local Authorities** manage **28.6%** of the total Natural Ramsar wetland area (439,830 ha), managing 29 sites. This implies that while the State handles the extremes (very small and very large sites), Local Authorities manage the mid sized natural sites. Thus, they act as connecting conduits between large anchor hubs and smaller spokes.

**Artificial Sites**

On similar lines as Natural sites, with **10 sites and 9.4% of the total area,** "**State Department"** is the dominant regulatory authority for Human-Made sites. Furthermore, **Local Authority** manages only **2 Human-Made sites** representing just **1.3% of the total area.** Alternatively, we also have a new category of governing authority - **Academic/Expert** - managing a lone site in **Ankasamudra Bird Conservation Reserve in Karnataka.**

### 
